Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-10-25 manuscript.tex

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, selection can remove genetically-linked neutral variation, resulting in reduced $N_{e}$ and the level of polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male fitness can also reduce Y-linked $N_{e}$, making it difficult infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked polymorphism during the early stages of plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). Strikingly, we find that nucleotide diversity on the Y is 40 fold lower than on the X, suggesting an extensive and rapid loss of diversity. We demonstrate that the magnitude of diversity loss is inconsistent with a purely neutral model based on the occurrence of female biased sex ratios in this species, or by reduced male $N_{e}$ caused by high variance in male fitness. Rather, using forward simulations, we show that Y diversity can be explained by purifying selection. Interestingly, whereas standard background selection theory predicted a diversity reduction n times what we observed, our simulations indicate that Y-linked variability can be explained by interference among a large number (>800kb) of loci subject to weak purifying selection (s). Given the recent origin of \textit{R. hastatulus} sex chromosomes, our study suggests that the low level of sequence diversity commonly observed on nonrecombining Y chromosomes might evolve relatively quickly after sex chromosomes originate, with selection against weakly deleterious mutations playing an important role during the earliest stages of their evolution.

@@

Several demographic factors are expected to modulate or accentuate differences in neutral polymorphism maintained on sex chromosomes \citep{ellegren2009}. These include population subdivision and sex-biased dispersal, deviations from a 1:1 breeding sex ratio, and high variance in reproductive success \citep{caballero1995,charlesworth2001,laporte2002,pool2007}. In theory, these processes all result in asymmetries in female vs male $N_{e}$, that could lead to different levels of neutral polymorphism on the sex chromosomes than are expected based on the autosomes (\citep{pool2007}). As evidence for this, coalescent simulations have shown that models of male-biased migration during human out-of-africa dispersal result in levels of X/A neutral diversity simular to what is observed \citep{keinan2009} (though it is worth noting that there is considerable variation in current estimates of X/A in humans, and other demographic scenarios have been suggested \citep{hammer2010,bustamante2009}). In addition, disproportionate losses of sex-linked compared to autosomal variation arising from population bottleneck events have been inferreed in chimpanzees and orangutans \citep{kaessmann2001,fischer2006}, the house mouse \textit{Mus musculus} \citep{baines2007}, and Drosophila \citep{andolfatto2001}, highlighting the importance of demography in affecting patterns of sex-linked and autosomal neutral variation.

In addition to demographic factors, evolutionary theory predicts that neutral variability on Y chromosomes with suppressed recombination can be significantly reduced by Hill-Robertson Interference (HRI), which refers to the reduction in $N_{e}$ of neutral sites that are genetically linked to sites experiencing selection. Two types of HRI include (i) selection on stronlgy benficial mutations ("selective sweeps") \citep{smith1974hitch,aquadro1994} and (ii) selection against strongly deleterious mutations
) \citep{charlesworth1996background,charlesworth1994effect} \citep{charlesworth1996CB,charlesworth2000degeneration}. In both processes, linkage  disequalibrium between selected and neutral sites results in a level of neutral variation below the level predicted in the absense of selection.

@@

%Despite widespread interest in determining the effects of recombination on patterns of genetic diversity \citep{ellegren2011,bachtrog2013NRG}, we still know very little about the influence of linked selection on patterns nucleotide polymorphism in younger sex chromosome systems, where \textit{de novo} recombination suppression evolved relatively recently \citep{charlesworth2016plant}. Given that background selection and selective sweeps are predicted to have the strongest effects during the earliest stages of Y-chromosome evolution, before Y-chromosomes have lost the majority of their genes , studying recently evolved sex chromosomes provides an important test of the temporal dynamics of the HR-effect.

>>>>

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, selection can remove genetically-linked neutral variation, resulting in reduced $N_{e}$ and the level of polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male fitness can also reduce Y-linked $N_{e}$, making it difficult to infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked polymorphism during the early stages of plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). Strikingly, we find that nucleotide diversity on the Y is 40-fold lower than on the X, suggesting an extensive reduction in diversity. We demonstrate that the magnitude of diversity loss is inconsistent with a purely neutral model that accounts for the occurrence of female-biased sex ratios in this species, or by reduced male $N_{e}$ caused by high variance in male fitness. Rather, using forward simulations and population genetic estimates of the distribution of fitness effects of deleterious mutations, we show that Y diversity levels can be explained by the action of background selection. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our study is consistent with the prediction that the early stages of Y chromsome degeneration are characterized by strong effects of selective interference experienced by a large number of selected sites. Although we do not exclude the possibility that positive selection has also reduced Y chromosome variability, our simulations suggest that the reduction in diversity arising from purifying selection is sufficient to explain our observed patters of neutral variation.

@@

Several demographic factors are expected to modulate or accentuate differences in neutral polymorphism maintained on sex chromosomes \citep{ellegren2009}. These include population subdivision and sex-biased dispersal, deviations from a 1:1 breeding sex ratio, and high sex-specific variance in reproductive success \citep{caballero1995,charlesworth2001,laporte2002,pool2007}. In theory, these processes all result in asymmetries in female vs male $N_{e}$, that could lead to different levels of neutral polymorphism on the sex chromosomes than are expected based on the autosomes (\citep{pool2007}). As evidence for this, coalescent simulations have shown that models of male-biased migration during human out-of-africa dispersal result in levels of X/A neutral diversity similar to what is observed \citep{keinan2009} (though it is worth noting that there is considerable variation in current estimates of X/A in humans, and other demographic scenarios have been suggested \citep{hammer2010,bustamante2009}). In addition, disproportionate losses of sex-linked compared to autosomal variation arising from population bottleneck events have been inferreed in chimpanzees and orangutans \citep{kaessmann2001,fischer2006}, the house mouse \textit{Mus musculus} \citep{baines2007}, and Drosophila \citep{andolfatto2001}, highlighting the importance of demography in affecting patterns of sex-linked and autosomal neutral variation.

In addition to demographic factors, evolutionary theory predicts that neutral variability on Y chromosomes with suppressed recombination can be significantly reduced by Hill-Robertson Interference (HRI), which refers to the reduction in $N_{e}$ of neutral sites that are genetically linked to sites experiencing selection. Two types of HRI include (i) selection on strongly benficial mutations ("selective sweeps") \citep{smith1974hitch,aquadro1994} and (ii) selection against strongly deleterious mutations
) \citep{charlesworth1996background,charlesworth1994effect} \citep{charlesworth1996CB,charlesworth2000degeneration}. In both processes, linkage  disequalibrium between selected and neutral sites results in a level of neutral variation below the level predicted in the absence of selection.

@@

Despite widespread interest in determining the effects of recombination on patterns of genetic diversity \citep{ellegren2011,bachtrog2013NRG}, we still know very little about the influence of linked selection on patterns nucleotide polymorphism in younger sex chromosome systems, where \textit{de novo} recombination suppression evolved relatively recently \citep{charlesworth2016plant}. Given that background selection and selective sweeps are predicted to have the strongest effects during the earliest stages of Y-chromosome evolution, before Y-chromosomes have lost the majority of their genes , studying recently evolved sex chromosomes provides an important test of the temporal dynamics of the HR-effect.
